Differences in score reliability do not explain meta-analytic heterogeneity in standardised effect sizes

### 2025-10-20 Manuscript_MD_ScoreReLiability.qmd
update manuscript - add abstracts

---
title: "Manuscript"
bibliography: ReLiability_references.bib
csl: styles/apa-6th-edition.csl
format: pdf
editor: 
  markdown: 
    wrap: 72
---

@@

# Liability or reliability? Exploring the role of effect size attenuation on meta-analytic heterogeneity

@@

@@

One source for this heterogeneity, which is not a typical moderator tied
to a specific theory, might be impairments in score reliability which
vary in their extent across replications. \[This is an argument put
forward repeatedly by original researchers and replicators alike, such
as Olsson-Collentine et al. [-@olsson2020] (do we know any more?).\]
Hunter and Schmidt [-@hunter2004] and Wiernik and Dahlke
[-@wiernik2020], discussing the role of score reliability for
standardized effect sizes, claim that differences in score reliability
should inflate, and therefore increase, heterogeneity. This implies that
standard corrections for imperfect score reliability, such as
attenuation correction procedures, should remove that inflation,
reducing heterogeneity in corrected effect sizes.

@@

If information concerning score reliability is available, it is possible
to correct the individual ES for its unreliability. This process is also
known as attenuation correction and, while not without its criticisms
[@muchinsky1996; @schmidt1996; @schennach2016], is a widespread practice
in Psychology (reference, alt.: repeatedly recommended + references).
@eq-dc describes a simple attenuation correction procedure.

@@

Re-analyses of the collaborative multi-site replication efforts have
uncovered differences in score reliability across replications of the
same phenomena [@shaw2020]. The claims made in W&D and H&S imply that
the ES heterogeneity uncovered in these projects therefore could
potentially be explained by these differences in score reliability.
Similarly, attempting to explain the substantial extent of ES
heterogeneity identified in psychological phenomena, both the original
authors as well as the multi-site replication authors claim that
differences in score reliability might be responsible (again, do we
actually know of any?).

@@

From the 21 phenomena that employed more than a single indicator, 19
made use of Likert-style items, where respondents would indicate their
agreement to some kind of statement. The scale-length varied from 3 to
10-points. The items measuring the remaining 2 phenomena were coded
dichotomously, as responses were right or wrong. While the majority of
designs make use of some form of control-treatment manipulation,
randomly assigning participants to a condition (19 phenomena), some
phenomena refer to the effect of pre-existing differences, e.g.
biological sex (2 phenomena). Subsequently, those phenomena need to be
identified where the meta-analytic mean effect size is statistically
significant different from zero. This will be done as a preliminary step
in this manuscript’s analysis procedure. Detailed information concerning
the different phenomena and how they were measured can be found at
\[osf-link\], while a brief summary can be found in @tbl-data.

@@

Subsequently, in order to assess whether ES heterogeneity was indeed
reduced by the attenuation correction procedure, the estimates of
heterogeneity in uncorrected ES and corrected ES are compared. Using
Cochran’s *Q*-test [@cochran1954], we assess whether the estimates of
heterogeneity are statistically significantly different from 0. Only for
those projects where we have sufficient confidence that the individual
estimates of heterogeneity are larger than zero can we actually begin to
interpret whether there was any change. \[Short paragraph discussing
that a test for difference in heterogeneity would be more informative?
Maybe move it to the discussion section - hint at it here?\]

@@

[Equations @eq-d], [-@eq-SEd], [-@eq-dc], and [-@eq-SEdc] were used to
generate estimates of standardized ES, both uncorrected and corrected,
with their corresponding standard errors. Generally, this leads to
larger (absolute) ES and larger standard errors after the correction. As
an example from the 21 selected phenomena, the results from this
procedure on replications of Nosek’s study on sex differences in
preferences for the subject of Arts is displayed in the forest plot in
@fig-forest. Here, grey dots represent the uncorrected ES for each
sample, while black dots represent the corrected ES in each sample. The
bars surrounding these dots show the respective 95%-Cis.

```{r, echo = FALSE, fig.height = 6}
#| label: fig-forest 
#| fig-cap: Forest Plot ML1 - replications of Nosek's studz on sex differences in subject preferences (Arts)
forest_plot_rel(ES_rma_df[which(ES_rma_df$MASC == "Nosek_Explicit_Art"),], agg_L[[which(MASC_names == "Nosek_Explicit_Art")]])
```

@@

gt::gt(read.csv(here("Tables/Mean_ES.csv")) %>% 
         arrange(MASC)) %>% 
  gt::cols_label(MASC = "MASC",
                 #mu_str_raw = gt::md("$\\hat{\\mu}_{\\delta_0}$ (SE)"),
                 mu_str_raw = gt::md("Mean $d$ (SE)"),
                 pval_raw = "p",
                 #mu_str_cor = gt::md("$\\hat{\\mu}_{\\delta}$ (SE)"),
                 mu_str_cor = gt::md("Mean $d_c$ (SE)"),
                 pval_cor = "p") %>% 
  gt::tab_options(table.width = gt::pct(100),
                  column_labels.font.weight = "bold") # %>% 
  # gt::cols_width(pval_raw ~ gt::px(40),
  #                pval_cor ~ gt::px(40))
```

*Note:* In this table *Mean* $d$ represents the meta-analytic estimate
of mean uncorrected $d$ and *Mean* $d_c$ represents the estimate of the
meta-analytic mean of corrected $d_c$.

@@

@tbl-tauES contains the results from performing an RG-MA on the
Cronbach's Alpha coefficients of the scales used to measure the 12
remaining phenomena. Additionally, it contains estimates of
heterogeneity for both uncorrected ES ($\delta_0$) and corrected ES
($\delta$), which are central to assess the research question posed in
section *Attenuation Correction*. Most importantly, @tbl-tauES
demonstrates that for all but one of the phenomena, the attenuation
correction procedure barely has an impact on the extent of ES
heterogeneity. However, while the heterogeneity for the phenomenon from
RRR9 (Behaviour) seems to be reduced by about a third of its initial
value, the ES heterogeneity for this phenomenon is rather small to begin
with. Using a significance test, we can not distinguish this value
confidently from zero. This implies that for the majority of phenomena,
differences in score reliability can not be made responsible for
heterogeneity in standardized effect sizes.

@@

gt::gt(tab_both  %>% 
  arrange(tab_hetES.MASC)) %>% 
  gt::cols_label(tab_hetES.MASC = "MASC",
                 mu_alpha_str = gt::md("$\\hat{\\mu}_{\\rho_{XX'}}$"),
                 tau_alpha = gt::md("$\\hat{\\tau}_{\\rho_{XX'}}$"),
                 QE_str = "QE (df)",
                 QEp = "p",
                 tau_raw = gt::md("$\\hat{\\tau}_{\\delta_0}$"),
                 QE_raw_str = "QE (df)",
                 QEp_raw = "p",
                 tau_raw = gt::md("$\\hat{\\tau}_{\\delta}$"),
                 QE_raw_str = "QE (df)",
                 QEp_raw = "p") %>% 
  gt::tab_options(table.width = gt::pct(100),
                  column_labels.font.weight = "bold") #%>% 
  # gt::cols_width(pval_raw ~ gt::px(40),
  #                pval_cor ~ gt::px(40))

@@

*Note:* $\hat{\mu}_{\rho_{XX'}}$ is the back-transformed meta-analytic
estimate of mean score reliability; $\tau_{\mu_{\rho_{XX'}}}$ is the
back-transformed estimate of heterogeneity in score reliability;
$\hat{\tau}_{\delta_0}$ represents the estimate of heterogeneity in
uncorrected $d$; $\hat{\tau}_{\delta}$ represents the estimate of the
heterogeneity in corrected $d_c$.

@@

Besides the direction of these differences, it is noteworthy that considering
differences in score reliability barely affects heterogeneity in standardized
effect sizes. [ausführen, oder kommt das noch einmal wieder?]

@@

Hunter & Schmidt have additionally derived an equation which
demonstrates how heterogeneity in score reliability affects
heterogeneity in ESs [-@hunter2004, p. XX]. @eq-HS implies that
heterogeneity in $\delta_0$ is essentially a function in which
heterogeneity and mean value of $\delta$ are weighted by mean score
reliability $\rho_{xx‘}$ and its heterogeneity.

@@

$R_1$ describes the relative size of mean true score variance, compared
to the mean observed score variance. This informs us, how much of the
mean total score variance can be attributed to mean true score variance.
Therefore, this metric is close to the average score reliability. Even
though formally $R_1$ and the meta-analytic mean score reliability
$\mu_{\rho_{xx'}}$ do not share not the same value, in practice we
observe little to no differences between the two parameters. [add to appendix]. Large
values indicate that correcting the individual score variances (or
standard deviations for that matter) using the attenuation correction
procedure should lead to very little change in ES, as most of the
variation observed were due to genuine differences in true scores. Small
values indicate the opposite, the observed score variance was due to a
larger amount of random error score variance, leading to a stronger
correction. The metric $R_2$ on the other hand describes in how far the attenuation
correction procedure has successfully reduced heterogeneity in score
variance $\tau^2_{\sigma^2_X}$.

@@

As long as there are substantial differences in the true score variance
across administrations, it is unlikely that an attenuation correction
can actually produce such a low $R_2$. Generally speaking, $R_1$ largely
depends on the relative measurement precision. If measurements were made
more precisely, average score reliability and $R_1$ would be higher. On
the other hand, $R_2$ depends on how different the measurement precision
was at the individual administrations, but also how different the
individual administrations sites are regarding their true score
variance. If the administration sites carry large differences in true
score variance $R_2$ is unlikely to be substantially smaller than $R_1$. [Na ja, eine Reliabilität von .9 bedeutet erstmal, dass es 9mal so viel wahre Varianz wie Fehlervarianz gibt. In diesem Fall müsste R2 unter 0.73 liegen. Das impliziert in der Heterogenität ein Verhältnis von ungefähr 3 : 1. Die 9mal größere wahre Varianz müsste also für weniger als ¾ der Heterogenität verantwortlich sein. In Deinem Textbeispiel (Reliabilität von .8) müsste die 4mal größere wahre Varianz für weniger als die Hälfte der Heterogenität verantwortlich sein. – Das erscheint mir beides per se ziemlich unplausibel… Also auch dann, wenn die „administration sites do NOT carry large differences in true score variance”. – In anderen Worten: Die Frage ist, ob die absolute Größe der Varianz überhaupt irgendeine Rolle spielt oder ob alles lediglich an den Verhältnissen hängt, die durch die Reliabilität vorgegeben sind. Ich habe dazu aber auch keine eindeutige Antwort… Wenn ich überzeugt wäre, dass sich *homogene* true score variances durch planvolle, kontrollierte Stichprobenziehung herstellen lassen, wäre das ein Gegenargument.]

@@

For the two exceptions in @tbl-tauES, those phenomena where the attenuation correction
procedure actually led to a decrease in ES heterogeneity (RRR9
(Behaviour) and ML5 - Shnabel (RPP)), we find that $R_2$ is larger than $R_1$.
This clearly contradicts our interpretation of @eq-ineq_int. As long as $R_2$ is
larger than $R_1^3$, we expect a correction procedure to increase effect 
size heterogeneity. However, we recommend to be cautious when interpreting both the change and
the direction of said change for these two projects. 
The significance test for the ES
heterogeneity was not significant, both before and after the correction.
Since either the heterogeneity was so small or the sample
size so low that the heterogeneity could not be statistically
distinguished from zero, most likely, the reduction in
heterogeneity is better explained by imprecise estimation than an actual
reduction of heterogeneity.

@@

@@

The analytical arguments, detailed in *Appendix A*, leading to
@eq-ineq_short, rest on the use of the delta-method. @eq-ineq_init
contains parameters of the score standard deviations’ distributions,
while the $R_1$ and $R_2$ metrics were designed to discuss score
variances. The delta-method rests on a number of assumptions, at least
one of which is likely violated in our analysis: the variance of the
variable to be transformed needs to be small [@cramer1999]. This entire
procedure discusses the influence of variance in standard deviations on
ES heterogeneity, introduced by differences in score reliability.
Therefore, assuming that this assumption holds seems naive. In the
*Appendix B*, we explore the archival data, to understand in how far
employing the delta-method on meta-analyses of standard deviations lead
to biased results than the meta-analyses on score variances we performed
using the Boot-Err technique. The differences identified are rather
miniscule for the majority of phenomena, at the third decimal. Only for
the phenomenon in PSACR002 (Behaviour) we find larger differences -
therefore, conclusions for this phenomenon concerning the relation
between $R_1$ and $R_2$ need to be drawn with caution. Generally, as the
majority of phenomena did not display any bias in this assessment and
since the delta-method is not actually used to generate estimates, we
believe its use to be defensible nonetheless.

Similarly, all equations following @eq-XYZ_cov rest on the assumption
that mean difference ($MD$) and true or total score variance
($\sigma^2_T$ and $\sigma^2_X$) are independent. While this assumption
is in line with assumptions underlying CTT, it is by no means guaranteed
that it holds in these data-sets. If this assumption was violated,
whether the relationship was positive or negative, indicated by a
positive or negative covariance in @eq-XYZ_cov, would most likely have
an impact on how strongly ES heterogeneity is affected by the
attenuation correction procedure, possibly even the direction of how it
is affected. However, currently we have no reason to assume that there
is a systematic relationship between mean difference and standard
deviation. As this relationship would also affect the distribution of ES
like Cohen’s d itself, strong violations of this assumption would go far
beyond invalidating the claims we derived in @eq-ineq_short. Most
likely, assumptions of the meta-analytic tests concerning the
meta-analytic mean size ES and the presence of ES heterogeneity would be
violated, making any research on change in ES heterogeneity due to
attenuation correction obsolete.

@@

### A - Rearranging inequality

Both metrics $R_1$ and $R_2$ make use of parameters of the distribution
of true and total score variance, not standard deviations. This choice
was made to facilitate the role of score reliability in the subsequent
discussion, as score reliability itself is defined by variances, not
standard deviations (see @eq-rel).

Using the metrics $R_1$ and $R_2$, defined in [Equations @eq-R1] and
[-@eq-R2], we attempt to understand under which circumstances the
inequality defined in @eq-ineq_init holds true. However, while the
metrics make use of mean and heterogeneity of the score variances,
@eq-ineq_init makes use of mean and heterogeneity of the standard
deviations. In order to incorporate the metrics, it is necessary to
reparameterise the equation. Using the delta method to approximate mean
value and heterogeneity of variance from those of the standard
deviations, we know that (ref):
$$\mu^2_{\sigma_X} \approx \mu_{\sigma^2_X} \quad \quad  and \quad \quad 
\tau^2_{\sigma_X} \approx \frac{\tau^2_{\sigma_X^2}}{4\mu_{\sigma^2_X}}$$ {#eq-delta_method}

While @eq-delta_method only explicitly contains parameters for the
observed score variance/standard deviation, the same can be done using
the true score variance/standard deviation. Using the approximates
defined in @eq-delta_method for observed and true score variance, we
arrive at a new inequality in @eq-ineq_CV.

@@

Secondly, as all terms contain $\mu_{\sigma^2_x}$ in the denominator, to
arrive at @eq-ineq_rearr2 we divided by $\mu_{\sigma^2_x}$.

@@

Taking $\frac{\tau_{\sigma^2_x}}{\mu_{\sigma^2_x}} = CV{\sigma^2_x}$
allows to simplify for @eq-ineq_rearr4.

@@

The circumstances under which the inequality described in @eq-ineq_fin
holds, are circumstances where the claims made in W&D and H&S are
directly contradicted, as in those cases the ES heterogeneity is larger
after applying an attenuation correction procedure. Concerning the terms
in @eq-ineq_fin, we know that all terms left-hand of the inequality,
outside of the brackets ($\tau^2_{MD}$, $\mu^2_{MD}$ and
$CV^2_{\sigma^2_X}$), are bound to be positive. For the inequality to
hold, we need the left-hand side of the equation to remain positive,
larger than zero. We can distill two scenarios, under which this
equation should hold: (a) one of the terms inside the brackets
($\frac{1}{R_1} - 1$ or $\frac{R_2}{R^3_1}  - 1$) is positive and large
enough, so that the second term not containing that same bracket is
positively dominated by the first term, meaning it is sufficiently large
that the positive term cancels out the negative term; or (b) both terms
inside the brackets need to be positive.

@@

However, it is crucial to note that violation of $R_1^3 \leq R_2$ is
just a necessary, but not sufficient condition for the inequality not to
hold anymore. Should $R_1^3 \leq R_2$ not hold, the negative term in
@eq-ineq_fin still needs to be sufficiently large, that it can
negatively dominate the positive term. That would mean that the
combination of parameters $\mu_{MD}$, $\tau_{MD}$ and $CV_{\sigma^2_x}$
need to take on values, that allow for such a scenario. Since we do not
want to make any claims regarding the unstandardized mean difference and
how it relates to the coefficient of variation of score variance, we can
not distill a scenario in which it is guaranteed, that @eq-ineq_fin is
violated. Therefore $R_1^3 \leq R_2$ only describes a condition, under
which it is guaranteed that @eq-ineq_fin holds. If $R_1^3 \leq R_2$ is
violated, it is a first necessary condition for the entire inequality to
be violated, but it is not sufficient as it still depends on other
parameters.

@@

@@

>>>>

---
title: "Differences in score reliability do not explain meta-analytic heterogeneity in standardised effect sizes"
bibliography: ReLiability_references.bib
csl: styles/apa-6th-edition.csl
format: docx
editor: 
  markdown: 
    wrap: 72
---

@@

#

@@

#### Abstract

Heterogeneity in standardized effect sizes, such as Cohen’s d, is informative because it indicates that a psychological phenomenon is not yet fully understood. Prior work by Hunter and Schmidt as well as Wiernik and Dahlke has argued that such heterogeneity partly reflects differences in measurement precision and that correcting for these differences using attenuation correction should therefore reduce heterogeneity. We reanalyze data from large-scale collaborative projects, including the ManyLabs studies, Registered Replication Reports, and the Psychological Science Accelerator, and demonstrate that this assumption does not hold universally. Treating standardized effect sizes as ratio variables, we offer an alternative account of how differences in measurement precision relate to observed heterogeneity. Across many psychological research contexts, score reliability tends to be high and measurement error relatively stable. Under these conditions, when samples differ substantially in their composition or diversity, attenuation correction procedures are more likely to increase rather than decrease heterogeneity. We conclude by discussing the limited relevance of random measurement error for explaining effect size heterogeneity in the context of convenience sampling.

#### Non-technical abstract
Researchers often find that studies on the same psychological effect report different results, known as differences in effect sizes. Such variation suggests that the phenomenon is not yet fully understood. Some have argued that part of this variation comes from differences in measurement precision—how reliably a study’s tools or tests measure what they intend to measure. Based on this reasoning, it has been suggested that adjusting or “correcting” for these differences should make results more consistent across studies. We revisited this assumption by looking at large collaborative projects in psychology, such as ManyLabs, the Registered Replication Reports, and the Psychological Science Accelerator. We found that the expected reduction in variation does not generally occur. In fact, when measurement precision is already quite high - as it usually is in psychology, where researchers carefully select reliable measures - but samples differ widely in who participates, such corrections can actually increase differences in reported effect sizes. In simpler terms, differences in how diverse the samples of participants are across studies (which comes about from convenience sampling) matters more for explaining variation in results than small differences in how precisely a measure captures a construct. We introduce a simple rule that identifies when correcting for measurement differences is likely to backfire, increasing rather than reducing heterogeneity. We conclude by discussing why random measurement error plays a smaller role in explaining effect size differences than the use of diverse, convenience-based samples.

@@

One source for this heterogeneity, which is not a typical moderator tied
to a specific theory, might be impairments in score reliability which
vary in their extent across replications [@olsson2020]. Hunter and
Schmidt [-@hunter2004] and Wiernik and Dahlke [-@wiernik2020],
discussing the role of score reliability for standardized effect sizes,
claim that differences in score reliability should inflate, and
therefore increase, heterogeneity. This implies that standard
corrections for imperfect score reliability, such as attenuation
correction procedures, should remove that inflation, reducing
heterogeneity in corrected effect sizes.

@@

If information concerning score reliability is available, it is possible
to correct the individual ES for its unreliability. This process is also
known as attenuation correction and, while not without its criticisms
[@muchinsky1996; @schmidt1996; @schennach2016], is a repeatedly
recommended practice in Psychology [@hunter2004, @wiernik2020] . @eq-dc
describes a simple attenuation correction procedure.

@@

Re-analyses of the collaborative multi-site replication efforts have
uncovered differences in score reliability across replications of the
same phenomena [@shaw2020]. The claims made in W&D and H&S imply that
the ES heterogeneity uncovered in these projects therefore could
potentially be explained by these differences in score reliability.
Similarly, in a re-analysis of multi-site replication projects,
Olsson-Collention et al. [-@olsson2020] discuss differences in score
reliability as a potential moderator for ES heterogeneity.

@@

From the 21 phenomena that employed more than a single indicator, 19
made use of Likert-style items, where respondents would indicate their
agreement to some kind of statement. The scale-length varied from 3 to
10-points. The items measuring the remaining 2 phenomena were coded
dichotomously, as responses were right or wrong. While the majority of
designs make use of some form of control-treatment manipulation,
randomly assigning participants to a condition (19 phenomena), some
phenomena refer to the effect of pre-existing differences, e.g.
biological sex (2 phenomena). Subsequently, those phenomena need to be
identified where the meta-analytic mean effect size is statistically
significant different from zero. This will be done as a preliminary step
in this manuscript’s analysis procedure. Detailed information concerning
the different phenomena and how they were measured can be found at
\[https://osf.io/g2fzq\], while a brief summary can be found in appendix C.

@@

Subsequently, in order to assess whether ES heterogeneity was indeed
reduced by the attenuation correction procedure, the estimates of
heterogeneity in uncorrected ES and corrected ES are compared. Using
Cochran’s *Q*-test [@cochran1954], we assess whether the estimates of
heterogeneity are statistically significantly different from 0. Only for
those projects where we have sufficient confidence that the individual
estimates of heterogeneity are larger than zero can we actually begin to
interpret whether there was any change.

@@

[Equations @eq-d], [-@eq-SEd], [-@eq-dc], and [-@eq-SEdc] were used to
generate estimates of standardized ES, both uncorrected and corrected,
with their corresponding standard errors. Generally, this leads to
larger (absolute) ES and larger standard errors after the correction. As
an example from the 21 selected phenomena, the results from this
procedure on replications of Nosek’s study on sex differences in
preferences for the subject of Arts is displayed in the forest plot in
@fig-forest. Here, grey dots represent the uncorrected ES for each
sample, while black dots represent the corrected ES in each sample. The
bars surrounding these dots show the respective 95%-CIs.

```{r, echo = FALSE, fig.height = 6}
#| label: fig-forest 
#| fig-cap: Forest Plot ML1 - replications of Nosek's study on sex differences in subject preferences (Arts)
forest_plot_rel(ES_rma_df[which(ES_rma_df$MASC == "Nosek_Explicit_Art"),], agg_L[[which(MASC_names == "Nosek_Explicit_Art")]])
```

*Note:* Grey dots represent uncorrected estimates of $d$, while black
dots represent corrected estimates of $d_c$. Similarly, grey bars
visualise the 95%-CI for uncorrected estimates, black bars the 95%-CIs
for corrected estimates. The grey diamond represents the meta-analytic
estimate of uncorrected ES $\mu_{d}$, its width represents the 95%-CI of
said estimate. The black diamond represents the same parameters for the
meta-analytic estimate of corrected ES $\mu_{d_c}$,.

@@

gt::gt(read.csv(here("Tables/Mean_ES.csv")) %>% 
         arrange(MASC)) %>% 
  gt::cols_label(MASC = "MASC",
                 mu_str_raw = gt::md("$\\hat{\\mu}_{d}$ (SE)"),
                 #mu_str_raw = gt::md("Mean $d$ (SE)"),
                 pval_raw = "p",
                 mu_str_cor = gt::md("$\\hat{\\mu}_{d_c}$ (SE)"),
                 #mu_str_cor = gt::md("Mean $d_c$ (SE)"),
                 pval_cor = "p") %>% 
  gt::tab_options(table.width = gt::pct(100),
                  column_labels.font.weight = "bold") # %>% 
  # gt::cols_width(pval_raw ~ gt::px(40),
  #                pval_cor ~ gt::px(40))
```

*Note:* In this table $\hat{\mu}_{d}$ represents the meta-analytic
estimate of uncorrected $d$ and $\hat{\mu}_{d_c}$ represents the
estimate of the meta-analytic mean of corrected $d_c$.

@@

@tbl-tauES contains the results from performing an RG-MA on the
Cronbach's Alpha coefficients of the scales used to measure the 12
remaining phenomena. Additionally, it contains estimates of
heterogeneity for both uncorrected ES ($d$) and corrected ES ($d_c$),
which are central to assess the research question posed in section
*Attenuation Correction*. Most importantly, @tbl-tauES demonstrates that
for all but one of the phenomena, the attenuation correction procedure
barely has an impact on the extent of ES heterogeneity. However, while
the heterogeneity for the phenomenon from RRR9 (Behaviour) seems to be
reduced by about a third of its initial value, the ES heterogeneity for
this phenomenon is rather small to begin with. Using a significance
test, we can not distinguish this value confidently from zero. This
implies that for the majority of phenomena, differences in score
reliability can not be made responsible for heterogeneity in
standardized effect sizes.

@@

gt::gt(tab_both  %>% 
  arrange(tab_hetES.MASC)) %>% 
  gt::cols_label(tab_hetES.MASC = "MASC",
                 mu_alpha_str = gt::md("$\\hat{\\mu}_{\\rho_{XX'}}$"),
                 tau_alpha = gt::md("$\\hat{\\tau}_{\\rho_{XX'}}$"),
                 QE_str = "QE (df)",
                 QEp = "p",
                 tau_raw = gt::md("$\\hat{\\tau}_{d}$"),
                 QE_raw_str = "QE (df)",
                 QEp_raw = "p",
                 tau_raw = gt::md("$\\hat{\\tau}_{d_c}$"),
                 QE_raw_str = "QE (df)",
                 QEp_raw = "p") %>% 
  gt::tab_options(table.width = gt::pct(100),
                  column_labels.font.weight = "bold") #%>% 
  # gt::cols_width(pval_raw ~ gt::px(40),
  #                pval_cor ~ gt::px(40))

@@

*Note:* $\hat{\mu}_{\rho_{XX'}}$ is the back-transformed meta-analytic
estimate of mean score reliability; $\tau_{\mu_{\rho_{XX'}}}$ is the
back-transformed estimate of heterogeneity in score reliability;
$\hat{\tau}_{d}$ represents the estimate of heterogeneity in uncorrected
$d$; $\hat{\tau}_{d_c}$ represents the estimate of the heterogeneity in
corrected $d_c$.

@@

Generally, however, it is noteworthy that removing differences in score
reliability only had a very little impact on ES heterogeneity. It
appears that, across the phenomena assessed in this project, the
heterogeneity before and after the correction procedure barely changes.
At least for the projects re-analysed here, differences in score
reliability do not affect ES heterogeneity as strongly as assumed.

@@

Hunter & Schmidt have additionally derived an equation which
demonstrates how heterogeneity in score reliability affects
heterogeneity in ESs [-@hunter2004, p. 309]. @eq-HS implies that
heterogeneity in $\delta_0$ is essentially a function in which
heterogeneity and mean value of $\delta$ are weighted by mean score
reliability $\rho_{xx‘}$ and its heterogeneity.

@@

$R_1$ describes the relative size of mean true score variance, compared
to the mean observed score variance. This informs us, how much of the
mean total score variance can be attributed to mean true score variance.
Therefore, this metric is close to the average score reliability. Large
values indicate that correcting the individual score variances (or
standard deviations for that matter) using the attenuation correction
procedure should lead to very little change in ES, as most of the
variation observed were due to genuine differences in true scores. Small
values indicate the opposite, the observed score variance was due to a
larger amount of random error score variance, leading to a stronger
correction. The metric $R_2$ on the other hand describes in how far the
attenuation correction procedure has successfully reduced heterogeneity
in score variance $\tau^2_{\sigma^2_X}$.

@@

As long as there are substantial differences in the true score variance
across administrations, it is unlikely that an attenuation correction
can actually produce such a low $R_2$. Generally speaking, $R_1$ largely
depends on the relative measurement precision. If measurements were made
more precisely, average score reliability and $R_1$ would be higher. On
the other hand, $R_2$ depends on how different the measurement precision
was at the individual administrations, but also how different the
individual administrations sites are regarding their true score
variance.

In our example of $R_1 = .8$, the true score variance is 4 times as large
as the error score variance. However, for the attenuation correction to
reduce ES heterogeneity, the smaller error score variance would need to
account for almost 50 % of heterogeneity in score variance. This asymmetry
grows even larger, as the average score reliability grows larger.
This demonstrated that the individual size of either
mean score variance component or its heterogeneity is not of importance.
Instead, the ratio of these components determine, whether heterogeneity
can decrease after an attenuation correction.

@@

For the two exceptions in @tbl-tauES, those phenomena where the
attenuation correction procedure actually led to a decrease in ES
heterogeneity (RRR9 (Behaviour) and ML5 - Shnabel (RPP)), we find that
$R_2$ is larger than $R_1$. This clearly contradicts our interpretation
of @eq-ineq_init. As long as $R_2$ is larger than $R_1^3$, we expect a
correction procedure to increase effect size heterogeneity. However, we
recommend to be cautious when interpreting both the change and the
direction of said change for these two projects. The significance test
for the ES heterogeneity was not significant, both before and after the
correction. Since either the heterogeneity was so small or the sample
size so low that the heterogeneity could not be statistically
distinguished from zero, most likely, the reduction in heterogeneity is
better explained by imprecise estimation than an actual reduction of
heterogeneity.

@@

*Note:* $\hat{\mu}_{\sigma^2_X}$ is the back-transformed meta-analytic
estimate of score variance; $\hat{\mu}_{\sigma^2_T}$ is the
back-transformed meta-analytic estimate of score variance;
$\tau_{\mu_{\sigma^2_X}}$ is the back-transformed estimate of
heterogeneity in score variance; $\hat{\tau}_{{\sigma^2_X}}$ represents
the estimate of heterogeneity in true score variance.

@@

The analytical arguments presented here, represented by @eq-ineq_fin,
hold as far as the use of the delta method does not lead to widely
erroneous results. In *Appendix C*, we compare estimates derived from
the delta-method and Boot-Err to assess in how far the delta-method
leads to biased estimates. Across all but one data-set, differences were
miniscule, leading us to believe that its use here was defensible. Only
for the PSACR002-data-set, we identify larger differences. While we
believe that $R_1$ and $R_2$ were estimated sufficiently well,
@eq-ineq_fin does not necessarily hold for how score reliability
heterogeneity affected ES heterogeneity here.

Similarly, all equations following @eq-XYZ_cov rest on the assumption
that mean difference ($MD$) and true or total score variance
($\sigma^2_T$ and $\sigma^2_X$) are independent. While this assumption
is in line with assumptions underlying CTT, it is by no means guaranteed
that it holds in these data-sets. If this assumption was violated, the
impact of an attenuation correction procedure on ES heterogeneity would
likely be affected. Whether this could lead to larger or lower ES
heterogeneity after the procedure than we currently assume depends on
whether the covariance in @eq-XYZ_cov is positive or negative. However,
generally we have no reason to assume that there is a systematic
relationship between mean difference and standard deviation. As this
relationship would also affect the distribution of ES like Cohen’s d
itself, strong violations of this assumption would go far beyond
invalidating the claims we derived in @eq-ineq_short. Most likely,
assumptions of the tests concerning the meta-analytic mean ES and the
presence of ES heterogeneity would be violated, making any research on
change in ES heterogeneity due to attenuation correction obsolete.

@@

### A - Rearranging the inequality

Both metrics $R_1$ and $R_2$ make use of parameters of the distributions
of true and total score variance, not standard deviations. This choice
was made to facilitate the discussion of score reliability, as score
reliability itself is defined by variances, not standard deviations (see
@eq-rel).

Using the metrics $R_1$ and $R_2$, defined in [Equations @eq-R1] and
[-@eq-R2], we attempt to understand under which circumstances the
inequality defined in @eq-ineq_init holds true. However, while the
metrics make use of mean and heterogeneity of the score variances,
@eq-ineq_init makes use of mean and heterogeneity of the standard
deviations. In order to incorporate the metrics, it is necessary to
reparameterise the equation. Using the delta method to approximate mean
value and heterogeneity of variance from those of the standard
deviations, we know that [@casella2002]:
$$\mu^2_{\sigma_X} \approx \mu_{\sigma^2_X} \quad \quad  and \quad \quad 
\tau^2_{\sigma_X} \approx \frac{\tau^2_{\sigma_X^2}}{4\mu_{\sigma^2_X}}$$ {#eq-delta_method}

While @eq-delta_method only contains parameters for the observed score
variance/standard deviation, the same can be done using the true score
variance/standard deviation. Using the approximates defined in
@eq-delta_method for observed and true score variance, we arrive at a
new inequality in @eq-ineq_CV.

@@

Secondly, as all terms contain $\mu_{\sigma^2_x}$ in the denominator, to
arrive at @eq-ineq_rearr2 we multiplied by $\mu_{\sigma^2_x}$.

@@

We define $CV{\sigma^2_x}$ as
$CV{\sigma^2_x} = \frac{\tau_{\sigma^2_x}}{\mu_{\sigma^2_x}}$, thereby
allowing to simplify for @eq-ineq_rearr4.

@@

The circumstances under which the inequality described in @eq-ineq_fin
holds, are circumstances where the claims made in W&D and H&S are
directly contradicted, as in those cases the ES heterogeneity is larger
after applying an attenuation correction procedure. Concerning the terms
in @eq-ineq_fin, we know that all terms left-hand of the inequality,
outside of the brackets ($\tau^2_{MD}$, $\mu^2_{MD}$ and
$CV^2_{\sigma^2_X}$), are bound to be positive. For the inequality to
hold, we need the left-hand side to remain positive, larger than zero.
We can distil two scenarios, under which this equation should hold: (a)
one of the terms inside the brackets ($\frac{1}{R_1} - 1$ or
$\frac{R_2}{R^3_1}  - 1$) is positive and large enough, so that the
second term not containing that same bracket is positively dominated by
the first term, meaning it is sufficiently large that the positive term
cancels out the negative term; or (b) both terms inside the brackets
need to be positive.

@@

However, it is crucial to note that violation of $R_1^3 \leq R_2$ is
just a necessary, but not sufficient condition for the inequality not to
hold anymore. Should $R_1^3 \leq R_2$ not hold, the negative term in
@eq-ineq_fin still needs to be sufficiently large, that it can
negatively dominate the positive term. That would mean that the
combination of parameters $\mu_{MD}$, $\tau_{MD}$ and $CV_{\sigma^2_x}$
need to take on values, that allow for such a scenario. Since we do not
want to make any claims regarding the unstandardised mean difference and
how it relates to the coefficient of variation of score variance, we can
not distil a scenario in which it is guaranteed, that @eq-ineq_fin is
violated. Therefore $R_1^3 \leq R_2$ only describes a condition, under
which it is guaranteed that @eq-ineq_fin holds. If $R_1^3 \leq R_2$ is
violated, it is a first necessary condition for the entire inequality to
be violated, but it is not sufficient as it still depends on other
parameters.

@@

The analytical arguments, detailed in *Appendix A*, leading to
@eq-ineq_short, rest on the use of the delta-method [@vandervaart2000,
@cramer1999, @casella2002]. @eq-ineq_init contains parameters of
distributions of the standard deviations of scores, while the $R_1$ and
$R_2$ metrics were designed to discuss score variances. The delta-method
rests on a number of assumptions, at least one of which is likely
violated in our analysis: the variance of the variable to be transformed
needs to be small [@cramer1999]. This entire procedure discusses the
influence of variance in standard deviations on ES heterogeneity,
introduced by differences in score reliability. Therefore, assuming that
this assumption holds seems naive. Here, we explore the archival data,
to understand in how far employing the delta-method on meta-analyses of
standard deviations lead to results biased more strongly than estimates
from the meta-analyses on score variances we performed using the
Boot-Err technique.

@@

### C - Summary of Phenomena, designs and measures
